---
title: "Reproducible Research: Peer Assessment 1"
author: "Sholeh Yasini"
date: 8 Jul 2017
output: 
  html_document:
    keep_md: true
---

```{r initialization}
library(ggplot2)
```

## Loading and preprocessing the data

```{r load_the_data}
activity <- unzip("activity.zip")
activity_data <- read.csv("activity.csv", sep = ",")
activity_data$date <- as.Date(activity_data$date)
```


## What is mean total number of steps taken per day?

1. Histogram of total number of steps per day

```{r mean-steps}
# sum the number of steps per day
sumDay <- aggregate(steps ~ date, data = activity_data, FUN = sum) 
# plot the histogram
hist(sumDay$steps, breaks = 10, col = "blue",
     main = "Total number of steps per day",
     xlab = "Steps per day")
```

2. The **mean** and **median** of the total number of steps taken each day:

```{r mean_total_steps}
mean(sumDay$steps, na.rm = TRUE)
median(sumDay$steps, na.rm = TRUE)
```



## What is the average daily activity pattern?

1. Time series plot of 5 minutes interval and the average number of steps 

```{r timeseries-plot}
minute <- activity_data$interval %% 100
hour <- activity_data$interval %/% 100
activity_data$newInt <- hour * 60 + minute

meanInt <- aggregate(steps ~ newInt, data = activity_data, FUN = "mean")

plot(meanInt$newInt, meanInt$steps, type = "l", col = "blue",
     xlab = "5-min interval", ylab = "average number of steps",
     main = "Timeseries plot of mean number of steps per interval")

```


## Imputing missing values

1. Calulate and report the total numner of missing values:

```{r total_missing_values}
missingval <- sum(is.na(activity_data$steps))
```

The total number of missing values are `r missingval`.

2. Devise a strategy for filling in all of the missing values in the dataset.

```{r replace missing value}
newdata <- activity_data
for (i in 1:nrow(newdata)){
    if(is.na(newdata$steps[i])){
        newdata$steps[i] <- meanInt$steps[meanInt$newInt == newdata$newInt[i]]
    }
}

s <- sum(is.na(newdata$steps))
```
The total number of missing values of the steps of the new dataset is `r s`.

3. The new dataset with no missing value is called **newdata**.

4. Hsitogram of total number of steps taken each day

```{r hist of newdata}
newSumDay <- aggregate(steps ~ date, data = newdata, FUN = "sum")

hist(newSumDay$steps, breaks = 10, col = "blue",
     main = "Total number of steps per day",
     xlab = "Steps per day")
```

The mean and median of total number of steps per day

```{r mean and median}
mean(newSumDay$steps)
median(newSumDay$steps)

```

While the mean is the same the median has a slight increase by considering the missing values. The total number of steps increases around the mean value.  


## Are there differences in activity patterns between weekdays and weekends?

```{r weekday and weekend patterns}
dates <- as.POSIXct(newdata$date)
weekday <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
newdata$weekday <- factor(weekdays(dates) %in% weekday, levels = c(FALSE, TRUE), labels = c("weekend","weekday"))

g <- ggplot(data = newdata, aes(x = newInt, y = steps)) + 
    stat_summary(fun.y = mean, geom = "line") + 
    facet_grid(. ~ weekday) + labs(x = "5 min interval", y = "average number of steps")
print(g)
```

Higher number of steps in weekends than weekdays except for the morning hours in weekdays.

